---
title: "Russell Project Exploratory data"
author: "Todd Curtis"
date: "10/31/2018"
geometry: margin=2cm
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Data project purpose###
The purpose if this data project is to take preprocessed spreadsheet of selected accident and incident data and put it into a form that could be anlayzed either directly with spreadsheet-based tools, or 
analyzed using a data analysis program such as **R**.

####Database description###
Each record of the database contains information for up to 51 variables. The database is contained in a spreadsheet, and each row or record represents an event involving a specific aircraft, as well as information that would describe the circumstances of the flight; the damage, injuries, or deaths that may have occurred during the flight; and the kind of accident or incident investigation conducted for the event. If an event involved multiple aircraft, there would be a separate row or record for each aircraft. 

A data dictionary describing these variables is available in the supplemental document *Proposed variables for database*.

While the database covers the period from January 1997 to October 2018, the period of interest in this study is the ten-year period from 2008 to 2017. 
```{r, echo=FALSE}
#Note that the `echo = FALSE` parameter was added to the code chunk to 
# prevent printing of the R code.

# Russell Project Exploratory data

# Purpose is to take preprocessed spreadsheet data and put it into a form 
# That can be analyzed using R

# ====== OVERVIEW OF DATABASE ===========
# There are 51 variables (column names) for the raw data. 
# A data dictionary describing these variables is available 
# in the supplemental document "Proposed variables for database"

# The type of data in the variables include:
#       Integer
#       Date (day-month-yyyy)
#       Factor (What is typically called a category variable)
#       Time (in UTC in four digit format)
#       Character

# ADMINISTRATIVE NOTES
# Note: To describe database at any point, use str(*name*)
# Note: To clear R workspace, use rm(list = ls())
# Note: Searching R help files - RSiteSearch("character string")
# Note: To clear console, use CTRL + L 


# ==== STATISTICS PACKAGES =====
# Just in case needed in future

options(repos = c(CRAN = "http://cran.rstudio.com"))
if("e1071" %in% rownames(installed.packages()) == FALSE) 
{install.packages("e1071")}
library(e1071)

if("rmarkdown" %in% rownames(installed.packages()) == FALSE) 
{install.packages("rmarkdown")}
library(rmarkdown)

# ==== DATA INPUT AND CLEANING ====

# READ DATA INTO DATA FRAME
# Import raw data (data files online in local directory AirSafe Consulting/Russell reports/event_database.csv)
events = NULL
events.raw = NULL
events.range = NULL
events.raw <- read.csv("event_database.csv", header = TRUE)

# Ensure that working data is in a data frame 
# Note: The raw data in file 'events.raw' will remain on standby
events = as.data.frame(events.raw)

# Ensure that working data has only complete rows
events <- events[complete.cases(events), ]


# Make everything start formatted as characters
events = as.data.frame(apply(events, 2, as.character)) 


# Using the function trimws removes leading and trailing space characters
events = as.data.frame(apply(events, 2, trimws))

# ENSURE Event.ID VARIABLE IS UNIQUE

# Variable designated identifier variable (Event.ID) variable
# is unique (number of unique values equal to the number of records)

# First step is to see if there is only one variable with unique values

unique.ndx = which(lapply(lapply(events,unique),length)==nrow(events))

# Hold off on printing this section in markdown document
# if (length(unique.ndx)==0){
#         print("No variable can serve as a unique identifier.")
# }
# 
# if (length(unique.ndx)==1)paste("Variable '", colnames(events)[unique.ndx],
#                          "' can serve as a unique identifier.", sep = "")
# 
# if (length(unique.ndx)>1){
#         print("The following variables that can serve as unique identifiers:")
#         colnames(events)[unique.ndx]
# }

# CONVERT VARIABLES TO THEIR APPROPRIATE TYPES
# There are several types of data in the data frame, including
#       integers, factors, character, and date. They will 
#       be transformed prior to any anaysis


# ==Convert integer variables==

integer.vars = c("Damage.Severity", "Aircraft.Cert.Year", "Production.Year", "Total.Crew", 
                 "Crew.Deaths", "Crew.Serious.Injuries", "Total.Passengers", 
                 "Passenger.Deaths", "Passenger.Serious.Injuries", 
                 "Other.Deaths", "Other.Serious.Injuries", "Total.Deaths", 
                 "Total.Serious.Injuries")

# Index of integer variables
integer.vars.ndx = which(colnames(events) %in% integer.vars)
         
# Convert these columns from factor to numeric

# Ensure integer variables are coerced to be numeric (will be integers)
for (i in 1:length(integer.vars.ndx)) {
        events[,integer.vars.ndx[i]] = as.numeric(as.character(events[,integer.vars.ndx[i]]))
}

# Identify missing or undefined data points for integer variables
# These integer values would have either 999 or 999 as the missing, unknown, or incompete integer values
integer.null.values = c(999,9999)

# Make missing or unknown integer data into NA data points
for (i in 1:length(integer.vars.ndx)) {
        events[which(events[,integer.vars.ndx[i]] %in% integer.null.values),integer.vars.ndx[i]] = NA
}

# Hold off on printing this section in markdown document

#Print out integer variables
# paste("Integer variables")
# colnames(events[integer.vars.ndx])


# ==Convert Factor variables==
factor.vars = c("Event.Type", "Operator.Category", "Aircraft.Type", "Flight.Phase", 
                "Flight.Purpose","Event.Country", "Event.IATA", "Origin.IATA", "Destination.IATA", 
                "Loss.Severity", "International.Flight", "Passenger.Flight",          
                "Scheduled.Flight", "Event.Special",     
                "Aircraft.Model", "Original.Operator","Engine.Type",
                "Investigation.Status")

# Index of factor variables
factor.vars.ndx = which(colnames(events) %in% factor.vars)


# Convert these columns to factors

# Ensure factor variables are coerced to be factors
for (i in 1:length(factor.vars.ndx)) {
        events[,factor.vars.ndx[i]] = as.factor(events[,factor.vars.ndx[i]])
}

# Identify missing or undefined data points for factor variables
# These factor values would have the following values as the missing, unknown, or incompete factor values
factor.null.values = c( "", "99", "9","999","9999", 
                       "X", "XX", "XXX", "XXXX",
                       "x", "xx", "xxx", "xxxx")


# Make missing or unknown factor data into NA data points
for (i in 1:length(factor.vars.ndx)) {
        events[which(events[,factor.vars.ndx[i]] %in% factor.null.values),factor.vars.ndx[i]] = NA
}

# Evaluating all database events
# Factor variables are categorical variables that can 
# be either numeric or string variables. 
# The variable definition document provides detailed 
# Description of both the variable and the categories 
# associated with each variable.


# Hold off on printing this section in markdown document
# Print factor variables
# paste("Factor variables")
#colnames(events[factor.vars.ndx])


# ==Convert Character variables==
# Note:Time variable 'Event.Time' is treated as a character so that it may be
#       reformatted later.

char.vars = c("Event.ID", "Event.Time", "Tail.Number", "Event.Location",
        "Origin", "Destination", "Flight.End", "Aircraft.Manufacturer", "Aircraft.Submodel",     
        "Registration.Country", "Serial.Number","Engine.Manufacturer", "Engine.Model", "Engine.Submodel",           
        "Operator.Name", "Operator.Country", "Investigating.Authority", "Probable.Cause", "Notes")

# Index of character variables
char.vars.ndx = which(colnames(events) %in% char.vars)

# Hold off on printing this section in markdown document
# Print character variables
# paste("Character variables")
# colnames(events[char.vars.ndx])

# Convert these columns to character

# Ensure character variables are coerced to be characters
for (i in 1:length(char.vars.ndx)) {
        events[,char.vars.ndx[i]] = as.character(events[,char.vars.ndx[i]])
}


# Identify missing or undefined data points for character variables
# These character values would have the following values as the missing, unknown, or incompete factor values
char.null.values = c("X", "XX", "XXX", "XXXX", "9999",
                       "x", "xx", "xxx", "xxxx")


# Make missing or unknown factor data into NA data points
for (i in 1:length(char.vars.ndx)) {
        events[which(events[,char.vars.ndx[i]] %in% char.null.values),char.vars.ndx[i]] = NA
}


# ==Convert Date Variable==

# There is only one date variable, Event.Date, with raw data format DD-Month-YYYY.
#       Will change format to R standard, so will turn it to character and
#       then to standard date format YYY-MM-DD.

events$Event.Date = as.character(events$Event.Date)
events$Event.Date = as.Date(events$Event.Date, "%d-%B-%Y")

# Hold off on printing this section in markdown document
# paste("Date variable is Event.Date")

# ==Convert Time Variable==

# The time variable is in the 24 hour clock format. CSV files save
#       it as an integer, so it is treated as such when it comes to
#       identifying a missing time value, which is coded as 9999. 

# The variable 'Event.Time' is in the 24 hour (HHMM) time format,  
#       but the leading zeros may have been eliminated during
#       initial uploading. The goal is to put it into the 
#       form HH:MM. Must do a combination of adding leading zeros 
#       and inserting a colon to get to HH:MM format

# Paste string "00:0 if one character long
one.long.ndx = which(nchar(events$Event.Time)==1)

events$Event.Time[one.long.ndx] = paste0("00:0",events$Event.Time[one.long.ndx])

# Add two leading zeros and colon if two characters long
# Becasue 'NA' is two characters long, need to 
#       find intersection of those not NA and those
#       two characters long

non.NA.values = which(!is.na(events$Event.Time))
two.long.all = which(nchar(events$Event.Time)==2)                     
two.long.ndx = intersect(non.NA.values, two.long.all)

events$Event.Time[two.long.ndx] = paste0("00:",events$Event.Time[two.long.ndx])

# Add one leading zero insert colon before last two characters if three characters long
three.long.ndx = which(nchar(events$Event.Time)==3)

events$Event.Time[three.long.ndx] = paste0("0",substr(events$Event.Time[three.long.ndx], 1,1),":",substr(events$Event.Time[three.long.ndx], 2,3))


# Insert colon after second character if time is four characters long (1000 to 2359)
four.long.ndx = which(nchar(events$Event.Time)==4)

events$Event.Time[four.long.ndx] = paste0(substr(events$Event.Time[four.long.ndx], 1,2),":",substr(events$Event.Time[four.long.ndx], 3,4))


# ====Post cleaning actions====
# After cleaning the database, including setting missing
# or unknown variables to NA, will summarize both the entire database and the 
# portion of the database that is of greatest interest.
# For the initial study, the is the ten-year period 2008-2017 inclusive.


```



```{r, echo=FALSE}

# === Summary of the study period ===

# Hold off on printing this in markdown document
# paste("Study period: 2008-2017")

# Create an index of records of aircraft events that fall in the study period
events.range.ndx = events$Event.Date>=as.Date("2008-01-01") & events$Event.Date<=as.Date("2017-12-31")
events.range = events[events.range.ndx,]
events.range.unique.ndx = which(substr(events.range$Event.ID, nchar(events.range$Event.ID)-1, nchar(events.range$Event.ID))=="01")

# Hold off on printing this in markdown document
# paste("Aircraft event date range for study period:", 
#      min(events.range$Event.Date),"to",
#      max(events.range$Event.Date))

# paste("Number of aircraft events during the study period -",nrow(events.range))
# 
# paste("Number of unique events during the study period -",length(events.range.unique.ndx))


# Table of events by year (during study period)

# Hold off on printing this in markdown document
# paste("Table of events by year (during study period)")
# table(as.integer(format(as.Date(events.range$Event.Date, format="%Y-%m-%d"),"%Y")))

# For events with and loss severity considered to be major
# (over $5M, ) there was a range range of percentage of arcraft value lost.

# Hold off on printing this in markdown document
# paste("Distribution of percent of aircraft damaged")
# table(events.range$Damage.Severity[which(events.range$Damage.Severity>0)])

# Hold off on printing this in markdown document
# paste("Distribution of percent of aircraft damaged for major events (over $5M losses)")
# table(events.range$Damage.Severity[which(events.range$Damage.Severity>0 & events.range$Loss.Severity=="M")])


# Summary of data from the factor and integer variables

# Variables that can be reasonably summarized include
# all factor and integer variables. Using the sorted 
# index values will result in summaries in the same
# order as in the spreadsheet that makes up the database
# The ordered index is sort(c(integer.vars.ndx,factor.vars.ndx))

# Hold off on printing this in markdown document
# paste("Summaries of data from factor and integer variables for study period")
# apply(events.range[,sort(c(integer.vars.ndx,factor.vars.ndx))],2,table, useNA = "always")

# Hold off on printing this in markdown document
# Can indirectly compute number of multiple events.
# Do a table of the last two Event.ID characters. 
# paste("Table of the last two Event.ID characters (multiple event indicator)")

# Hold off on printing this in markdown document
# table(substr(events.range$Event.ID, nchar(events.range$Event.ID)-1,nchar(events.range$Event.ID)))
```
###High level summary###

Because each record incudes detailed informaiton about each event, including time, date, and location of each occurrence, and also information that categorizes each event by flight phase, cause, and level of economic loss, there are many kinds of questions that can be asked and answered. As a first step, it may help to have a summary of the kind of data that is available:

* Study period: 2008-2017

* Aircraft event date range during study period: `r format(min(events.range$Event.Date))` to `r format(max(events.range$Event.Date))`

* Number of aircraft events during the study period: `r format(nrow(events.range), big.mark=",")`

* Number of unique events during the study period: `r format(length(events.range.unique.ndx), big.mark=",")`


###Summaries using graphs###

One can use a basic graph to summarize the data and to see general patterns.

```{r, echo=FALSE}

# === BAR PLOT OF Years ===
# Bar plot of events by year
# Will only use unique events

year.vector = as.numeric(format(events.range$Event.Date[events.range.unique.ndx],'%Y'))

barplot(table(year.vector),
        main = "Unique events by year (2008-2017)",
        xlab = "Year",
        ylab = "Unique events",
        col = "dodgerblue")

# ==========
```


```{r, echo=FALSE}

# === BAR PLOT OF Months ===
# Bar plot of events by month and ensuring the order of months
# Will only use unique events

month.vector = months(events$Event.Date[events.range.unique.ndx], abbreviate = TRUE)
month.vector = factor(month.vector,levels=c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"), ordered=TRUE)

barplot(table(month.vector),
        main = "Unique events by month (2008-2017)",
        xlab = "Month",
        ylab = "Unique events",
        col = "dodgerblue")
# ==========
```


```{r, echo=FALSE}

# === BAR PLOT OF DAYS ===
# Bar plot of events by day of the week and ensuring order of days
# Will only use unique events

day.vector = weekdays(events.range$Event.Date[events.range.unique.ndx], abbreviate = TRUE)
day.vector = factor(day.vector,levels=c("Sun","Mon","Tue","Wed","Thu","Fri","Sat"), ordered=TRUE)

barplot(table(day.vector),
        main = "Unique events by weekday (2008-2017)",
        xlab = "Day",
        ylab = "Unique events",
        col = "dodgerblue")
# ==========
```


```{r, echo=FALSE}

# === BAR PLOT OF HOURS ===
# Using the 24-hour clock, so hours range from 0 to 23
# For the vector of hours, convert the first two characters of the 
# Event.Time variable to numeric form
# Will only use unique events

hour.vector = as.numeric(substr(events.range$Event.Time[events.range.unique.ndx],1,2))

barplot(table(hour.vector),
        main = "Unique events by hour (2008-2017)",
        xlab = "Hour",
        ylab = "Unique events",
        col = "dodgerblue")
```

###Using heatmaps to discover patterns###
One can use a basic data table to illustrate the relationship between two variables. The example below shows how the number of unique events varies by the day of the week and the month of the year. By using only a table, it may be hard to see subtle underlying patters:

```{r, echo=FALSE}

# Table of raw values
# Vector names changed to look better in table
Day = day.vector
Month = month.vector
table(Day,Month)
```

One can use a basic table a heat map to visually depict the information as was done with raw data, and by using scaling one can also show if either particular days of the week or particular months of the year have more unique events. 

The maps below are scaled three ways:

####1. No scaling#### 
This is a visual representation of the raw data table shown above. Darkest cells represent higher numbers of unique events for a particular combination of day of the week and month of the year. The darkest cells correspond indicate that this combination of month and day had more unique events than lighter colored cells.

```{r, echo=FALSE}

# Heat maps will use a color pallette that goes from white for lowest to dark blue for the highest value
 palette = colorRampPalette(c('#ffffff','#0000ff'))(64)
 
heatmap(table(day.vector,month.vector),Rowv=NA, Colv=NA,revC=TRUE,
        scale="none", col = palette, margins=c(6,6),
        main="Unique events by month and day of the week")
```

####2. Scaled by day####
Darker columns highlight the months with consistently
higher numbers of unique events reports compared to other months. Scaling by row (the day) will highlight the month of the year with a relatively high or low number of unique events. In other words, for each row (day of the week), it will rank the month from the most unique events (darkest) to the fewest (lightest). If a particular month of the year is consistently more likely to have events, that entire column (month of the year) will be in general darker for most days of the week. If that month of the year is less likely to have events, it will be relatively lighter each day of the week.

```{r, echo=FALSE}
# Heat maps will use a color pallette that goes from white for lowest to dark blue for the highest value
 palette = colorRampPalette(c('#ffffff','#0000ff'))(64)
 
heatmap(table(day.vector,month.vector),Rowv=NA, Colv=NA,revC=TRUE,
        scale="row", col = palette, margins=c(6,6),
        main="Unique events by month and day scaled by day")
```

####3. Scaled by month#### 
Darker rows highlight the days with consistently
higher numbers of unique events compared to other days of the week. Scaling by column (the month) will highlight the day of the week with a relatively high or low number of unique events. In other words, for each column (month), it will rank the days from the most unique events (darkest) to the fewest (lightest). If a particular day of the week is consistently more likely to have events, that entire row (day of the week) will be in general darker for most months. If that day of the week is less likely to have events, it will be relatively lighter for most months of the year.

```{r, echo=FALSE}
# Heat maps will use a color pallette that goes from white for lowest to dark blue for the highest value
 palette = colorRampPalette(c('#ffffff','#0000ff'))(64)
 
heatmap(table(day.vector,month.vector),Rowv=NA, Colv=NA,revC=TRUE,
        scale="column", col = palette, margins=c(6,6),
        main="Unique events by month and day scaled by month")
```


